TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-03-11 paper/draft.tex
restructure & add text to intro

\newcommand{\apogee}{\textsl{APOGEE}}
\newcommand{\thecannon}{\textsl{The Cannon}}
\newcommand{\cannon}{\textsl{Cannon}}
\newcommand{\aspcap}{\textsl{ASPCAP}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\zmass}{\textsl{2MASS}}
\newcommand{\sdss}{\textsl{SDSS}}
\newcommand{\lamost}{\textsl{LAMOST}}
\newcommand{\sdssiv}{\textsl{SDSS-IV}}
\newcommand{\tess}{\textsl{TESS}}
\newcommand{\teff}{$T_{\rm eff}$}
\newcommand{\feh}{[Fe/H]}

@@

Very low mass stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}. Relevance to exoplanet studies, and other reasons for scientific interest...

Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements. Citations to any other large M dwarf catalogs that have

The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy. What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?

Define what parameters/classifications are important for M dwarfs.

State why fundamental physical parameters are dificult to measure for M dwarfs. When temperatures and metallicities for a star are unavailable, it is often useful to classify a star in terms of spectral type. Define spectral type (this is a classification that is based on how a spectra looks, something to do with TiO band strengths, and it correlates with the temperature of a star).

Why is there a problem labeling M dwarfs (and especially for high res \apogee\ spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled. Several approaches to to determining physical parameters and the limitations:

\begin{enumerate}
\item model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013})--models don't accurately reproduce spectra plus unquantified systematic errors;

\item model narrow wavelegth region where models are good, look at specific features -- however we want to leverage the full wavelength range and resolution of our data.

\item using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters
\end{enumerate}

@@

\subsection{Radii}
% Using empirically calibrated relations from \citealt{Mann:2015} we determine radii... some diagnostics...
TBD

@@

TBD
% As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures. To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\textsl{APASS}; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by \textsl{APASS}, and $JHK$ photometries from \apogee. Figure \ref{fig:teff_comparisons} shows \cannon\ vs. photometric temperatures on the right, and \aspcap\ vs. photometric temperatures on the left for each of the 6,257 sources, colored by their respective spectroscopic metallicities.

@@

% We find \aspcap\ overestimates \teff\ on average by $130-170$K, while \thecannon\ underestimates \teff\ by $\sim25$K . We also see the scatter is much higher for the \aspcap\ temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).

@@

% While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.

\subsection{Isochrone Comparisons}
TBD

@@

%%==========================================================================================
\section{Discussion} \label{sec:discussion}
TBD

% \begin{enumerate}
% \item[-] Brief summary of what we did and found

% \item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.

% \item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}

% \item[-] Some discussion of whether the derivatives are sensible or surprising. \color{gcolor}{HOGG: I'm not sure what to say here}\color{black}

% \item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}

% \item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}

% \end{enumerate}

% Future work:
% \begin{enumerate}
% \item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}

% \item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
% \end{enumerate}

>>>>

\newcommand{\apogee}{\textsl{APOGEE}}
\newcommand{\thecannon}{\textsl{The Cannon}}
\newcommand{\cannon}{\textsl{Cannon}}
\newcommand{\aspcap}{\textsl{ASPCAP}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\zmass}{\textsl{2MASS}}
\newcommand{\sdss}{\textsl{SDSS}}
\newcommand{\kepler}{\textsl{Kepler}}
\newcommand{\lamost}{\textsl{LAMOST}}
\newcommand{\sdssiv}{\textsl{SDSS-IV}}
\newcommand{\tess}{\textsl{TESS}}
\newcommand{\msun}{$M_{\odot}$ \,}
\newcommand{\rsun}{$R_{\odot}$ \,}
\newcommand{\teff}{$T_{\rm eff}$}
\newcommand{\feh}{[Fe/H]}

@@

Low-mass stars (with masses $M_{*} < 0.7\,$\msun and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and XX of the population by mass. With nuclear fusion timescales of $\tau>10^{12}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.

Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations (e.g. \kepler, \tess), and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts. For these reasons, M dwarfs are the primary candidates for exoplanet detection searches, making detailed and precise knowledge of M dwarf chemical compositions key to constraining the atmospheric conditions of potentially habitable exoplanets through exoplanet spectra obtained through transit spectroscopy.

Precise measurements of fundamental stellar parameters (including temperatures, masses, radii, and abundances) are key to studying M dwarfs on both galactic and planetary scales. Methods for determining these parameters... temperatures (models, photometry), masses (dynamical)

Why is there a problem determining M dwarf metallicities (and especially for high res \apogee\ spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.

Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements. 
The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.

@@

\subsection{Masses \& Radii}
Masses and radii for our test sample are estimated using photometric empirical relations from \citealt{Mann:2019} and \citealt{Mann:2015} respectively.

@@

As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures. To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\textsl{APASS}; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by \textsl{APASS}, and $JHK$ photometries from \apogee. Figure \ref{fig:teff_comparisons} shows \cannon\ vs. photometric temperatures on the right, and \aspcap\ vs. photometric temperatures on the left for each of the 6,257 sources, colored by their respective spectroscopic metallicities.

@@

We find \aspcap\ overestimates \teff\ on average by $130-170$K, while \thecannon\ underestimates \teff\ by $\sim25$K. We also see the scatter is much higher for the \aspcap\ temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).

@@

While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.

\subsection{Isochrone Comparisons}

@@

%%==========================================================================================
\section{Discussion} \label{sec:discussion}

\begin{enumerate}
\item[-] Brief summary of what we did and found

\item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.

\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}

\item[-] Some discussion of whether the derivatives are sensible or surprising. \color{gcolor}{HOGG: I'm not sure what to say here}\color{black}

\item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}

\item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}

\end{enumerate}

Future work:
\begin{enumerate}
\item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}

\item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
\end{enumerate}
